Composition-climate paper

### 2015-07-07 Composition_Ecology.Rmd
Added new figures to help the `md` file render online.
Added code block to get `pandoc.table` to run properly.

Some text edits in the Introduction, adding references &cetera.

---
title: "Composition-climate paper"
author: "Simon Goring *et al*."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
pdf_document:
pandoc_args: "-H margins.sty"
bibliography: composition.bib
csl: harvard1.csl
---

@@

@@

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# Load climate data and generate the pre-sett and modern bounds:
#  Pre-settlement climate: 1900 - 1930
#  Modern climate: 1960 - 1990

@@

Over the past 200 years centennial scale climate variability and land use change have interacted to transform the forests of the northeastern United States. There are clear indications that modern forests, as represented by the Forest and Inventory Analysis show greater homogeneity and significantly different species associations than pre-settlement forests [@goring2015preset]. This shift has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change.

If modern species distributions have been affected by land use change in a random fashion then we should expect little effect on the resulting predictions.  In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie and savanna to closed forest, in an approixmately west to east direction, and from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.  Goring et al [@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, but did not examine how these shifts might affect the relationships between climate and distribution for individual taxa.

We use gridded climate data products and estimates of pre-settlement vegetation to develop climate-vegetation relationships for X major tree taxa in the upper Midwestern United States to understand how climate and land use change have interacted since Euro-American settlement to affect species distributions in the region, and to examine the possible implications of these shifts to future estimates of species resillience to climate change across the region.

Methods
===============
We pair pre-settlement vegetation data from the Public Land Survey System aggregated to a 64km^2^ grid [@goring2015] with annually resolved climate data for the region.  PLSS vegetation data contains no uncertainty, however it is spatially discontinuous, particularly in the south where digitization of PLSS records has occured at a slower rate.  Additionally, data in the northeastern United States [@thompson2013four] from town Proprietor Surveys is recorded arially, with no variance within township polygons.  To accomodate these different data sources a conditional autoregressive model was applied to the data, providing the ability to obtain uncertainty estimates across the domain [@paciorek2015comp].

@@

```
**Figure 1**. *Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.  Low temperatures in the 1970s result in lower T_max_ during the modern era, but climate change has resulted in higher T_min_ values than the early-century normals, along with increasing A_ppt_.  PDSI shows strong coherence with A_ppt_ in this region.*

@@

#  This takes a bit of time.
loss_plots <- loss_plot()

```
**Figure 1**.  *Gain loss and continuous presence for PLSS and FIA tree taxa.*

```{r densities}

@@

```
**Figure** *Changes in the overall climate space in the upper midwestern United States. The climate densities largely overlap, although it is possible to see the shifts in precipitation, mean and minium temperatures, where FIA data (pink) has higher values), while the decline in maximum temperatures is also visible, largely at the very highest values.*

@@

```
**Figure X**. *For each of the four variables examined we see*

>>>>

---
title: "Composition-climate paper"
author: "Simon Goring *et al*."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  md_document:
    variant: markdown
  pdf_document:
    pandoc_args: "-V geometry:vmargin=1in -V geometry:hmargin=1in"
pandoc_args: "-H margins.sty"
bibliography: goringetal_references.bib
csl: ecology.csl
---

@@

<!--

```{r setup-knitr, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  comment = " ",
  error = TRUE,
  eval = TRUE,
  cache = TRUE,
  tidy = TRUE

)

library("pander")
library("plyr")
panderOptions('table.style', 'rmarkdown')
```
-->

@@

```{r, echo = FALSE, warning = FALSE, message = FALSE, message=FALSE}
# Load climate data and generate the pre-sett and modern bounds:
#  Pre-settlement climate: 1900 - 1930
#  Modern climate: 1960 - 1990

@@

Over the past 200 years centennial scale climate variability and land use change have interacted to transform the forests of the northeastern United States. There are clear indications that modern forests, as represented by the Forest and Inventory Analysis show greater homogeneity and significantly different species associations than pre-settlement forests [@goring2015preset;@schulte2007homogenization]. This shift has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change [@iverson2013tree;@iverson1998predicting].  One of the most significant changes to forest tree distributions in the upper midwest is the loss of extensive cover through land use change [@rhemtulla2009legacies], particularly the .

```{r}

natural[[2]] + 
  geom_tile(data = land_use, aes(x = x, y = y, fill = class)) +
  scale_fill_manual(values = c('#ff7f00', '#e41a1c', '#33a02c', '#377eb8'))

```
**Figure 1.** *Patterns of land use change in the upper Midwestern United States. Data from the NLCD [@jin2013comprehensive].*

If modern species distributions have been affected by land use change in a random fashion then we should expect little effect on the resulting predictions since, by and large, me might expect the climate envelope encompased by the species to remain largely unchanged.  In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie/savanna to closed forest, in an approixmately west to east direction, and one from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.  Goring et al [@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, but did not examine how these shifts might affect the relationships between climate and distribution for individual taxa. Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests, we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.  Conversely, species in the north should show little change if we expect that pre-settlement trees show similar regeneration patterns (with respect to simple presence/absence) following widespread logging.

We use gridded climate data products and estimates of pre-settlement vegetation to develop climate-vegetation relationships for 15 major tree taxa in the upper Midwestern United States to understand how climate and land use change have interacted since Euro-American settlement to affect species distributions in the region, and to examine the possible implications of these shifts to future estimates of species resillience to climate change across the region.

Methods
===============
We pair pre-settlement vegetation data from the Public Land Survey System aggregated to a 64km^2^ grid [@goring2015preset] with annually resolved climate data for the region.  PLSS vegetation data contains no uncertainty, however it is spatially discontinuous, particularly in the south where digitization of PLSS records has occured at a slower rate.  Additionally, data in the northeastern United States [@thompson2013four] from town Proprietor Surveys is recorded arially, with no variance within township polygons.  To accomodate these different data sources a conditional autoregressive model was applied to the data, providing the ability to obtain uncertainty estimates across the domain [@paciorek2015comp].

@@

```
![](figures/clim_plot.tiff)
**Figure 1**. *Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.  Low temperatures in the 1970s result in lower T_max_ during the modern era, but climate change has resulted in higher T_min_ values than the early-century normals, along with increasing A_ppt_.  PDSI shows strong coherence with A_ppt_ in this region.*

@@

#  This takes a bit of time.
loss_melt <- loss_plot()

loss_table <- as.data.frame.matrix(table(loss_melt$variable, loss_melt$value))
loss_table$Gain <- loss_table$Gain / (loss_table$Loss + loss_table$Presence)
loss_table[,2:3] <- tran(loss_table[,2:3], method = 'proportion')

```
![](figures/loss_plots.tiff)

**Figure 2**.  *Gain loss and continuous presence for PLSS and FIA tree taxa. Continuous presence is indicated by dark green.  For most taxa it is the central parts of their distribution that show continuous presence.  Few taxa show novel presence (purple), although taxa such as poplar, ash and basswood do show gains of over 10%.  Most taxa show significant losses in distribution (orange).*

**Table 1.** *Gain and loss of the various forest tree types of interest.*
```{r, results=as.is}
pandoc.table(loss_table, justify = "left", style = 'simple')
```

```{r densities, warning=FALSE, message=FALSE, echo=FALSE}

@@

**Figure 3** *Changes in the overall climate space in the upper midwestern United States. The climate densities largely overlap, although it is possible to see the shifts in precipitation, mean and minium temperatures, where FIA data (pink) has higher values), while the decline in maximum temperatures is also visible, largely at the very highest values.*

@@

hell_plot[[1]]

**Figure X**. *For each of the four variables examined we see...*
